---
title: 'Bios 6301: Assignment 8'
author: Max Rohde
output: tint::tintPdf
---

```{r setup, include=FALSE}
library(tint)
library(tidyverse)
library(glue)

# invalidate cache when the package version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
options(htmltools.dir.version = FALSE)
```

## Question 1 ##

Install the `readxl` package and run the following

```{r}
fn <- 'icd10.xlsx'
if(file.access(fn, mode = 4) == -1) {
    url <- "https://www.cdc.gov/nhsn/xls/icd10-pcs-pcm-nhsn-opc.xlsx"
    download.file(url, destfile = fn, mode = 'wb')
}
dat <- readxl::read_excel(fn, sheet = 2)
```

1. Show the class of `dat`. (1 point)

```{r}
class(dat)
```


2. Show the methods available for objects of the given class (if there are multiple classes, show methods for all classes). (3 points)

```{r}
methods(class = c("tbl_df"))
```
```{r}
methods(class = c("tbl"))
```


```{r}
methods(class = c("data.frame"))
```


3. If you call `print(dat)`, what print method is being dispatched? (1 point)

```{r}
sloop::s3_dispatch(print(dat))

# We see that print.tbl_df is being dispatched
```

4. Set the class of `dat` to be a data.frame. (1 point)

```{r}
class(dat) <- "data.frame"
```


5. If you call `print(dat)` again, what print method is being dispatched? (1 point)

```{r}
sloop::s3_dispatch(print(dat))

# We see that print.tbl is being dispatched
```


Define a new generic function `nUnique` with the code below.

```{r}
nUnique <- function(x) {
    UseMethod('nUnique')
}
```

6. Write a default method for `nUnique` to count the number of unique values in an element. (2 points)

```{r}
nUnique.default <- function(x){
  length(unique(x))
}
```


7. Check your function (2 points)

```{r}
# should return 26 
nUnique(letters)

# should return 10 (probably)
nUnique(sample(10, 100, replace = TRUE)) 
```

8. Write a data.frame method for `nUnique` to operate on data.frame objects.
This version should return counts for each column in a data.frame. (2 points)

```{r}
nUnique.data.frame <- function(x){
  map(dat, ~nUnique(.x))
}
```


9. Check your function (2 points)

```{r}
nUnique(dat)
```

## Question 2 ##

Programming with classes.  The following function will generate random patient information.

```{r}
makePatient <- function() {
  vowel <- grep("[aeiou]", letters)
  cons <- grep("[^aeiou]", letters)
  name <- paste(sample(LETTERS[cons], 1),
                sample(letters[vowel], 1),
                sample(letters[cons], 1), sep = "")
  gender <- factor(sample(0:1, 1),
                   levels = 0:1,
                   labels = c("female", "male"))
  dob <- as.Date(sample(7500, 1), origin = "1970-01-01")
  n <- sample(6, 1)
  doa <- as.Date(sample(1500, n), origin = "2010-01-01")
  pulse <- round(rnorm(n, 80, 10))
  temp <- round(rnorm(n, 98.4, 0.3), 2)
  fluid <- round(runif(n), 2)
  list(name, gender, dob, doa, pulse, temp, fluid)
}
```

1. Create an S3 class `medicalRecord` for objects that are a list with the named elements `name`, `gender`, `date_of_birth`, `date_of_admission`, `pulse`, `temperature`, `fluid_intake`. Note that an individual patient may have multiple measurements for some measurements.  Set the RNG seed to `8` and create a medical record by taking the output of `makePatient`.  Print the medical record, and print the class of the medical record. (5 points)

```{r}
create_record <- function() {
  record <- makePatient()
  names(record) <- c("name",
                     "gender",
                     "date_of_birth",
                     "date_of_admission",
                     "pulse",
                     "temperature",
                     "fluid_intake")
  
  class(record) <- "medicalRecord"

  return(record)
}

set.seed(8)
record <- create_record()
```

```{r}
print(record)
```

```{r}
print(class(record))
```


2. Write a `medicalRecord` method for the generic function `mean`, which returns averages for pulse, temperature and fluids. Also write a `medicalRecord` method for `print`, which employs some nice formatting, perhaps arranging measurements by date, and `plot`, that generates a composite plot of measurements over time.  Call each function for the medical record created in part 1. (5 points)

```{r}
mean.medicalRecord <- function(x) {
  pulse_mean <- mean(x[["pulse"]])
  temperature_mean <- mean(x[["temperature"]])
  fluids_mean <- mean(x[["fluid_intake"]])

  measurements <- list(pulse_mean = pulse_mean,
                       temperature_mean = temperature_mean,
                       fluids_mean = fluids_mean)
  return(measurements)
}

mean(record)
```

```{r}
print.medicalRecord <- function(x) {
  print(glue("Name: {x$name}"))

  print(glue("Gender: {x$gender}"))

  print(glue("DOB: {x$date_of_birth}"))
  
  
  df <- tibble(date_of_admission = x$date_of_admission,
               pulse = x$pulse,
               temperature= x$temperature,
               fluid_intake = x$fluid_intake)
  
  df %>% arrange(date_of_admission) -> df
  
  print.data.frame(df)
}

print(record)
```

```{r}
plot.medicalRecord <- function(x) {
  plot(x$date_of_admission,
       x$pulse,
       type = "l",
       main = "Pulse")
  
  plot(x$date_of_admission,
       x$temperature,
       type = "l",
       main = "Temperature")
  
  plot(x$date_of_admission,
       x$fluid_intake,
       type = "l",
       main = "Fluid Intake")
}

plot(record)
```


3. Create a further class for a cohort (group) of patients, and write methods for `mean` and `print` which, when applied to a cohort, apply mean or print to each patient contained in the cohort. Hint: think of this as a "container" for patients.  Reset the RNG seed to 8 and create a cohort of ten patients, then show the output for `mean` and `print`. (5 points)

```{r}
set.seed(8)

# Create a cohort of 10 patients
cohort <- map(1:10, ~create_record())
class(cohort) <- "cohort"

mean.cohort <- function(x) {
  means <- map(x, ~mean(.x))
  return(means)
}

mean(cohort)
```

```{r}
print.cohort <- function(x) {
  walk(x, ~print(.x))
}

print(cohort)
```


