---
title: 'Bios 6301: Assignment 2'
author: "Max Rohde"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(kableExtra) # For making tables look nice
```

1. **Working with data** In the `datasets` folder on the course GitHub repo, you will find a file called `cancer.csv`, which is a dataset in comma-separated values (csv) format. This is a large cancer incidence dataset that summarizes the incidence of different cancers for various subgroups. (18 points)

    1. Load the data set into R and make it a data frame called `cancer.df`. (2 points)
    
```{r,  message=FALSE}
# Read in the data
cancer.df <- read_csv('cancer.csv')
```

    2. Determine the number of rows and columns in the data frame. (2)
    
```{r}
# Get the dimensions, ROWS x COLUMNS
dim(cancer.df)
```
There are 42120 rows and 8 columns in this data frame.

    3. Extract the names of the columns in `cancer.df`. (2)
    
```{r}
# Get the column names
names(cancer.df)
```

    4. Report the value of the 3000th row in column 6. (2)
    
```{r}
# Get the value of the 6th column of row 3000
# [[row, column]]
cancer.df[[3000,6]]
```
The value is 350.69

    5. Report the contents of the 172nd row. (2)
    
```{r}
# [ROW 172, all columns]
cancer.df[172,] %>%
  kbl(caption = "The 172nd row") %>%
  kable_classic(full_width = F, html_font = "Source Sans Pro")

# another approach
cancer.df %>% slice(172) %>%
  kbl(caption = "The 172nd row (tidyverse method)") %>%
  kable_classic(full_width = F, html_font = "Source Sans Pro")
```
    
    6. Create a new column that is the incidence *rate* (per 100,000) for each row. The incidence rate is the `(number of cases)/(population at risk)`, which in this case means `(number of cases)/(population at risk) * 100,000`. (3)
    
```{r}
# Calculate incidence rate
cancer.df <- mutate(cancer.df, inc_rate = (incidence / population) * 1e5)

# Check the results
head(cancer.df) %>%
  kbl(caption = "Added incidence rate") %>%
  kable_classic(full_width = F, html_font = "Source Sans Pro")
```

    7. How many subgroups (rows) have a zero incidence rate? (2)
```{r}
cancer.df %>%
  filter(inc_rate==0) %>%  # Filter to rows with zero incidence rate
  nrow() # Calculate number of rows
```
23191 subgroups have a zero incidence rate.
    

    8. Find the subgroup with the highest incidence rate.(3)
    
```{r}
cancer.df %>%
  arrange(desc(inc_rate)) %>% # Sort the data frame from highest incidence rate to lowest
  head(1) %>% # Take the first row
  kbl(caption = "Row with highest incidence rate") %>%
  kable_classic(full_width = F, html_font = "Source Sans Pro")
```
  The subgroup shown above has the highest incidence rate.

2. **Data types** (10 points)

    1. Create the following vector: `x <- c("5","12","7")`. Which of the following commands will produce an error message? For each command, Either explain why they should be errors, or explain the non-erroneous result. (4 points)

            max(x)
            sort(x)
            sum(x)
            
            
```{r}
x <- c("5","12","7")

# The max() command compares character strings in order from first to last character.
# Because 7 > 5 > 1, 7 is returned.
max(x)
```
            
            
```{r}
# To obtain the standard result (treating each element as a number)
# we can cast to integer first
x %>% as.integer() %>% max()
```
            
```{r}
# For the same reasons as above, the vector will be sorted as 
# 12, 5, 7 because only the first character is compared
# and 1 < 5 < 7
sort(x)
```
            
```{r}
# The usual result after conversion
x %>% as.integer() %>% sort()
```
```{r}
# sum() will return an error because it has no rules defined
# for adding together two character vectors

#sum(x)
```
```{r}
# we must first convert to an integer
x %>% as.integer() %>% sum()
```


    2. For the next two commands, either explain their results, or why they should produce errors. (3 points)

            y <- c("5",7,12)
            y[2] + y[3]
            
```{r}
# This code produces an error because when creating the vector `y`, 
# all of the elements are coerced to characters so when the 2nd
# line is run, we are adding "7" + "12". This is undefined
# because R does not have addition defined for characters
# (unlike a language like Python where they would be concatenated)

# y <- c("5",7,12)
# y[2] + y[3]

```
            
    3. For the next two commands, either explain their results, or why they should produce errors. (3 points)

            z <- data.frame(z1="5",z2=7,z3=12)
            z[1,2] + z[1,3]
            
```{r}
z <- data.frame(z1="5",z2=7,z3=12)
z[1,2] + z[1,3]

# This code works as expected, adding 7 + 12 = 19, unlike the last problem, 
# because while vectors can only have one type, data frames can have each column
# be a different type. Thus, 7 and 12 are allowed to be integers, and exist
# in the data frame with "5" because they are in their own columns, and each
# column is a vector of a single type.
```
            

3. **Data structures** Give R expressions that return the following matrices and vectors (*i.e.* do not construct them manually). (3 points each, 12 total)

    1. $(1,2,3,4,5,6,7,8,7,6,5,4,3,2,1)$
    
```{r}
count_to <- function(x) c(1:(x-1),x,(x-1):1)

count_to(8)
```
    
    
    2. $(1,2,2,3,3,3,4,4,4,4,5,5,5,5,5)$
    
```{r}
rep(1:5, times=1:5)
```
    
    3. $\begin{pmatrix}
      0 & 1 & 1  \\
      1 & 0 & 1  \\
      1 & 1 & 0  \\
    \end{pmatrix}$
    
```{r}
# diag(3) creates a 3x3 identity matrix
# then subtract 1 and take the absolute value to obtain the desired output

abs(diag(3) - 1)
```
    

    4. $\begin{pmatrix}
      1 & 2 & 3 & 4 \\
      1 & 4 & 9 & 16 \\
      1 & 8 & 27 & 64  \\
      1 & 16 & 81 & 256 \\
      1 & 32 & 243 & 1024  \\
    \end{pmatrix}$

```{r}
# First we create a list of numbers 1:5
m <- rep(1:4, 5)

# Then put into a matrix
m <- matrix(m, nrow=5, byrow = TRUE) 

# Then raise the first row to the first power, second row to second power...etc
m^(1:5)
```

4. **Basic programming** (10 points)

    1. Let $h(x,n)=1+x+x^2+\ldots+x^n = \sum_{i=0}^n x^i$. Write an R program to calculate $h(x,n)$ using a `for` loop. As an example, use `x = 5` and `n = 2`. (5 points)
    
```{r}
h <- function(x,n){
  sum <- 0
  for (i in 0:n){
    # Add each term of the polynomial to the sum
    sum <- sum + x^i
  }
  sum
}

h(5,2)
```
  
    1. If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23. Write an R program to perform the following calculations. (5 points)

        1. Find the sum of all the multiples of 3 or 5 below 1,000. (3, [euler1])
        
```{r}
div3 <- ((1:999) %% 3) == 0 # Boolean array of elements divisible by 3
div5 <- ((1:999) %% 5) == 0 # Boolean array of elements divisible by 5
div3_5 <- div3 | div5 # Divisible by 3 OR 5

(1:999)[div3_5] %>% sum() # Get the relevant elements by boolean indexing, then sum.
```
        

        1. Find the sum of all the multiples of 4 or 7 below 1,000,000. (2)
        
```{r}
# Same approach as last problem

div4 <- ((1:10^6-1) %% 4) == 0
div7 <- ((1:10^6-1) %% 7) == 0
div4_7 <- div4 | div7

(1:10^6-1)[div4_7] %>% sum()
```
    
    1. Each new term in the Fibonacci sequence is generated by adding the previous two terms. By starting with 1 and 2, the first 10 terms will be $(1, 2, 3, 5, 8, 13, 21, 34, 55, 89)$. Write an R program to calculate the sum of the first 15 even-valued terms. (5 bonus points, [euler2])
    
```{r}
# Stores the Fibonacci sequence
x <- numeric()

# Stores only the even terms
evens <- numeric()

# Define initial parameters
x[1] <- 1
x[2] <- 1

# We start the loop at 3, because the first two numbers are already defined
i <- 3

# We loop until we found 15 even fibonacci numbers
while(length(evens) < 15)
{
  # The new value is the sum of the previous two in the sequence
  new <- x[i - 2] + x[i - 1]
  x[i] <- new
  
  # If the new value is even, append it to the list of even numbers
  if((new %% 2) == 0) evens <- append(evens, new)

  # Increment the counter to move on to the next fibonacci number
  i <- i+1
}

# Sum the 15 even fibonacci numbers
s <- sum(evens)
```

```
The sum is `r format(s, scientific=FALSE)`
```

Some problems taken or inspired by projecteuler.

[euler1]: https://projecteuler.net/problem=1
[euler2]: https://projecteuler.net/problem=2
