---
title: 'Bios 6301: Assignment 7'
author: "Max Rohde"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
```


### Question 1 ###

**21 points**

Use the following code to generate data for patients with repeated measures of A1C (a test for levels of blood glucose).

```{r}
genData <- function(n) {
    if(exists(".Random.seed", envir = .GlobalEnv)) {
        save.seed <- get(".Random.seed", envir= .GlobalEnv)
        on.exit(assign(".Random.seed", save.seed, envir = .GlobalEnv))
    } else {
        on.exit(rm(".Random.seed", envir = .GlobalEnv))
    }
    set.seed(n)
    subj <- ceiling(n / 10)
    id <- sample(subj, n, replace=TRUE)
    times <- as.integer(difftime(as.POSIXct("2005-01-01"), as.POSIXct("2000-01-01"), units='secs'))
    dt <- as.POSIXct(sample(times, n), origin='2000-01-01')
    mu <- runif(subj, 4, 10)
    a1c <- unsplit(mapply(rnorm, tabulate(id), mu, SIMPLIFY=FALSE), id)
    data.frame(id, dt, a1c)
}
x <- genData(500)
```

Perform the following manipulations: (3 points each)

1. Order the data set by `id` and `dt`.

```{r}
# Change name and convert to tibble
df <- tibble(x)

# Order by `id` and `dt`
df %>%
  arrange(id, dt) -> df

df %>%
  head(10) %>%
  kableExtra::kable(booktabs=TRUE)
```

2. For each `id`, determine if there is more than a one year gap in between observations.  Add a new row at the one year mark, with the `a1c` value set to missing.  A two year gap would require two new rows, and so forth.

```{r}
# Create a column of difference in years between each timepoint
df %>%
  group_by(id) %>%
  mutate(difference = c(dt %>% diff() %>% time_length(unit = "year"),0)) %>%
  ungroup() -> df

# Filter only to differences greater than 1 year, and take the floor()
df %>%
  filter(difference >= 1) %>%
  mutate(difference = floor(difference)) -> diffs

# Remove the differences column
df <- df %>% select(-difference)

# Iterate through the differences table
# Add in the missing years
for(i in 1:nrow(diffs)){
  for(j in 1:diffs$difference[i]){
    df <- df %>% add_row(id = diffs$id[i],
                         dt = diffs$dt[i] + years(j),
                         a1c = NA)
  }
}

# Order by `id` and `dt`
df %>%
  arrange(id, dt) -> df

# View first 50 rows
df %>%
  head(50) %>%
  kableExtra::kable(booktabs=TRUE)
```

3. Create a new column `visit`.  For each `id`, add the visit number.  This should be 1 to `n` where `n` is the number of observations for an individual.  This should include the observations created with missing a1c values.

```{r}
df %>%
  group_by(id) %>%
  mutate(visit = 1:length(id)) -> df

df %>%
  head(20) %>%
  kableExtra::kable(booktabs=TRUE)
```

4. For each `id`, replace missing values with the mean `a1c` value for that individual.

```{r}
df %>%
  group_by(id) %>%
  mutate(a1c = ifelse(is.na(a1c), mean(a1c,na.rm=TRUE),a1c)) -> df
```

5. Print mean `a1c` for each `id`.

```{r}
df %>%
  group_by(id) %>%
  summarise(mean_a1c = mean(a1c, na.rm = TRUE)) %>%
  kableExtra::kable(booktabs=TRUE)
```

6. Print total number of visits for each `id`.

```{r}
df %>%
  group_by(id) %>%
  summarise(num_visit = max(visit)) %>%
  kableExtra::kable(booktabs=TRUE)
```

7. Print the observations for `id = 15`.
```{r}
df %>%
  filter(id == 15) %>%
  kableExtra::kable(booktabs=TRUE)
```




### Question 2 ###

**16 points**

Install the `lexicon` package. Load the `sw_fry_1000` vector, which contains 1,000 common words.

```{r}
data('sw_fry_1000', package = 'lexicon')
head(sw_fry_1000)
```

1. Remove all non-alphabetical characters and make all characters lowercase. Save the result as `a`.

```{r}
sw_fry_1000 %>%
  str_replace_all("\\W", "") %>%
  str_to_lower() -> a

# Examine the differences
sw_fry_1000[sw_fry_1000 != a]
a[sw_fry_1000 != a]
```

Use vector `a` for the following questions. (2 points each)

2. How many words contain the string "ar"?

```{r}
str_detect(a, "ar") %>% sum()

# We see that 64 words contain the string "ar"
```

3. Find a six-letter word that starts with "l" and ends with "r".

```{r}
a[str_detect(a, "^l.{4}r$")]

# The only match is "letter"
```

4. Return all words that start with "col" or end with "eck".

```{r}
a[str_detect(a, "(^col|eck$)")]
```

5. Find the number of words that contain 4 or more adjacent consonants. Assume "y" is always a consonant.

```{r}
str_detect(a, "[^aeiou]{4,}") %>% sum()

# 8 words fit this description
```

6. Return all words with a "q" that isn't followed by a "ui".

```{r}
a[str_detect(a, "q(?!ui)")]
```

7. Find all words that contain a "k" followed by another letter. Run the `table` command on the first character following the first "k" of each word.

```{r}
a[str_detect(a, "k.")]

str_match(a, "k(.)")[,2] %>% table()
```

8. Remove all vowels. How many character strings are found exactly once?

```{r}
no_vowel <- str_replace_all(a, "[aeiou]","")

(table(no_vowel) == 1) %>% sum()

# There are 581 strings found exactly once

```

### Question 3 ###

**3 points**

The first argument to most functions that fit linear models are formulas.  The following example defines the response variable `death` and allows the model to incorporate all other variables as terms. `.` is used to mean all columns not otherwise in the formula.

```{r}
url <- "https://github.com/couthcommander/Bios6301/raw/master/datasets/haart.csv"
haart_df <- read.csv(url)[,c('death','weight','hemoglobin','cd4baseline')]
coef(summary(glm(death ~ ., data=haart_df, family=binomial(logit))))
```

Now imagine running the above several times, but with a different response and data set each time.  Here's a function:

```{r}
myfun <- function(dat, response) {
  form <- as.formula(response ~ .)
  coef(summary(glm(form, data=dat, family=binomial(logit))))
}
```

Unfortunately, it doesn't work. `tryCatch` is "catching" the error so that this file can be knit to PDF.

```{r}
tryCatch(myfun(haart_df, death), error = function(e) e)
```

What do you think is going on?  Consider using `debug` to trace the problem.

#### Answer

The issue is that when `death` is passed into the function, R tries to interpret it as an object coming from the global namespace, where it doesn't exist, since it only exists in relation to the data frame.

**5 bonus points**

Create a working function.

```{r}
# Here is a working version using reformulate()
myfun <- function(dat, response) {
  form <- reformulate(".", deparse(substitute(response)))
  coef(summary(glm(form, data=dat, family=binomial(logit))))
}

myfun(haart_df, death)
```
